Evaluate robustness example

### 2021-10-25 notebook/_posts/2021-10-18-evaluate-robustness-example/evaluate-robustness-example.Rmd
change output from html to distill
Edit and render analysis
- Add reference and information about the source data
- Add 'Data setup' section with code to download and clean the source
  data
- Add session info
- Adjust the appearance of the summary plot
- Make minor text edits
- Run analysis with 100 iterations and render output to
  Distill-formatted html page

---
title: "Evaluate robustness example"
description: |
  Evaluate the robustness of an estimate and inference to bias.
author:
  - name: Amy D. Willis
date: 10-18-2021
output: html_document
---

@@

We can investigate the sensitivity of any analysis to taxonomic bias by simulating possible efficiency vectors, calibrating our observed MGS measurements by these efficiencies, and rerunning our analysis on the simulated and calibrated data. Specifically, we can take the observed proportion of each species in our dataset $\hat{\text{prop}}_{i}(a)$, and "undo" the efficiencies to obtain the calibrated proportions
\begin{align}
  (\#eq:calibrate-props)
  \text{calibrated prop}_{i}(a)
  &= \left( \frac{\hat{\text{prop}}_{i}(a)}{\text{efficiency}_{i}} \right) \bigg / \sum_{i'} \left( \frac{\hat{\text{prop}}_{i'}(a)}{\text{efficiency}_{i'}} \right),
\end{align}
and the calibrated read counts
\begin{align}
  (\#eq:calibrate-reads)
  \text{reads}_{i}(a)
  &= \text{total reads}(a) \times \text{calibrated prop}_{i}(a).
\end{align}
We can then rerun our data analysis using the calibrated read counts in place of our original data. If we do this repeatedly, each time using different realizations of the efficiency vector, we can investigate how the results of our analysis change. If the change is minimal, this suggests that our results are robust to unequal detection efficiencies in the different species. The converse would suggest that we need to be careful interpreting our results.

## Example: relative abundance of Gardnerella

For example, suppose we are interested in investigating if the genus Gardnerella is more abundant in the vaginal microbiome of women who give birth preterm, and we investigate this model by running `corncob` using preterm birth as a covariate in the mean and dispersion model. We could investigate the robustness of our findings using the following structure:

@@

First, we load the relevant libraries and our data

```{r}
# library(here)
library(tidyverse)
library(magrittr)
library(speedyseq)
library(corncob)
# devtools::install_github("jacobbien/simulator")
library(simulator)
cal <- readRDS("callahan2017repl_phyloseq_clean.rds")
```

@@

Next, we specify how we use our data to construct our estimates. We are using `bbdml` from the library `corncob` to model the relative abundance of ASV 8 (Gardnerella) using preterm birth as a covariate to model the mean and dispersion, but this is only one type of analysis that we might be interested in.

@@

Now that we have all of the pieces, we can run them together as follows. Note that we are only going to run 4 iterations for now, but we have commented out code below showing you how to run 100 draws across 4 cores:

```{r}
our_sim <- new_simulation("UndoGardnerella", "Undo Gardnerella relative abundance") %>%
  generate_model(undo_efficiencies,
                 ps = cal %>%
                   filter_sample_data(cohort == "Stanford") %>%
                   tax_glom("genus") %>%
                   as_tibble,
                 sig_sq_e = as.list(seq(0, 5, length.out = 6)),
                 vary_along = "sig_sq_e") %>%
  simulate_from_model(nsim = 4, index = 1) %>%
  # simulate_from_model(nsim = 25, index = 1:4) %>%
  run_method(list(corncob_Gardnerella)) %>%
  # run_method(list(corncob_Gardnerella),
  #            parallel = list(socket_names = 4, libraries = c("corncob", "speedyseq"))) %>%
  evaluate(list(betahat_fit, pvalue_fit))
```

Here's how the estimated coefficient on preterm birth changes under different draws from a distribution given by $\text{efficiency}_{i}} \overset{iid}{\sim} LogNormal(0, \sigma^2_e)$, and how the p-value for testing the null hypothesis that the coefficient on preterm birth is zero:

```{r}
ev_df <- our_sim %>% evals %>% as.data.frame
model_df <- our_sim %>% model %>% as.data.frame
ev_with_model_params <- dplyr::right_join(model_df, ev_df, by = c("name" = "Model")) %>% tibble
f_names <- list('p-value' = "p-value",
                'beta-hat' = expression(hat(beta)))
f_labeller <- function(variable, value){return(f_names[value])}
ev_with_model_params %>%
  pivot_longer(cols = betahat:pvalue, names_to="eval") %>%
  mutate(eval = ifelse(eval == "pvalue", "p-value", "beta-hat")) %>%
  ggplot(aes(x = sig_sq_e, group = sig_sq_e, y = value)) +
  geom_boxplot() +
  geom_abline(aes(slope = slope, intercept = intercept), color = "blue",
              tibble(eval = c("p-value", "beta-hat"), slope = c(0,0), intercept = c(0.05, 0))) +
  theme_bw() +
  xlab(expression(sigma[e]^2)) +
  ggtitle("Corncob results for testing\nH0: Mean Gardnerella relative abundance is equal for preterm & full-term") +
  facet_wrap(~eval, scales="free_y", nrow = 2, labeller = f_labeller) +
  ylab("")
```

@@

A major advantage of this approach is that any distribution of efficiency vectors can be investigated. For example, we could specifically investigate how Gardnerella being low- or high-efficiency impacts our results. We could also investigate more sophisticated distributions for the efficiencies, such as correlated efficiencies across phylogeny.

>>>>

---
title: "Evaluate robustness example"
description: |
  Evaluate the robustness of an estimate and inference to bias.
author:
  - name: Amy D. Willis
categories:
  - bias sensitivity
  - ref:callahan2017repl
date: 10-18-2021
bibliography: ../../../main.bib
output:
  distill::distill_article:
    self_contained: false
    toc: true
    dev: svg
---

@@

We can investigate the sensitivity of any analysis to taxonomic bias by simulating possible efficiency vectors, calibrating our observed MGS measurements by these efficiencies, and rerunning our analysis on the simulated and calibrated data. Specifically, we can take the observed proportion of each species in our dataset $\hat{\text{prop}}_{i}(a)$, and "undo" the efficiencies to obtain the calibrated proportions
\begin{align}
  \text{calibrated prop}_{i}(a)
  &= \left( \frac{\hat{\text{prop}}_{i}(a)}{\text{efficiency}_{i}} \right) \bigg / \sum_{i'} \left( \frac{\hat{\text{prop}}_{i'}(a)}{\text{efficiency}_{i'}} \right),
\end{align}
and the calibrated read counts
\begin{align}
  \text{reads}_{i}(a)
  &= \text{total reads}(a) \times \text{calibrated prop}_{i}(a).
\end{align}
We can then rerun our data analysis using the calibrated read counts in place of our original data. If we do this repeatedly, each time using different realizations of the efficiency vector, we can investigate how the results of our analysis change. If the change is minimal, this suggests that our results are robust to unequal detection efficiencies in the different species. The converse would suggest that we need to be careful interpreting our results.

## Example: relative abundance of _Gardnerella_

We illustrate this method using vaginal microbiome data from pregnant women published in a study by @callahan2017repl seeking to replicate associations of specific genera (_Lactobacillus_, _Gardnerella_, and _Ureaplasma_) with preterm birth in two cohorts.
For example, suppose we are interested in investigating if the genus _Gardnerella_ is more abundant in the vaginal microbiome of women who give birth preterm, and we investigate this model by running `corncob` using preterm birth as a covariate in the mean and dispersion model. We could investigate the robustness of our findings using the following structure:

@@

First, we load the relevant libraries and our data (see [Data setup]),

```{r}
# library(here)
library(tidyverse)
library(magrittr)
library(speedyseq)
library(corncob)
# devtools::install_github("jacobbien/simulator")
library(simulator)
cal <- readRDS(
  here::here("notebook/_data", "callahan2017repl", "callahan2017repl_phyloseq_clean.rds")
)
```

@@

Next, we specify how we use our data to construct our estimates. We are using `bbdml` from the library `corncob` to model the relative abundance of ASV 8 (_Gardnerella_) using preterm birth as a covariate to model the mean and dispersion, but this is only one type of analysis that we might be interested in.

@@

Now that we have all of the pieces, we can run them together as follows.
The below code runs 100 simulations split across 4 cores; 
for testing, the commented code runs just 4 iterations.

```{r}
our_sim <- new_simulation("UndoGardnerella", "Undo Gardnerella relative abundance") %>%
  generate_model(undo_efficiencies,
                 ps = cal %>%
                   filter_sample_data(cohort == "Stanford") %>%
                   tax_glom("genus") %>%
                   as_tibble,
                 sig_sq_e = as.list(seq(0, 5, length.out = 6)),
                 vary_along = "sig_sq_e") %>%
  # simulate_from_model(nsim = 4, index = 1) %>%
  # run_method(list(corncob_Gardnerella)) %>%
  simulate_from_model(nsim = 25, index = 1:4) %>%
  run_method(list(corncob_Gardnerella),
             parallel = list(socket_names = 4, libraries = c("corncob", "speedyseq"))) %>%
  evaluate(list(betahat_fit, pvalue_fit))
```

Here's how the estimated coefficient on preterm birth changes under different draws from a distribution given by $\text{efficiency}_{i} \overset{iid}{\sim} \text{LogNormal}(0, \sigma^2_e)$, and how the p-value for testing the null hypothesis that the coefficient on preterm birth is zero:

```{r}
ev_df <- our_sim %>% evals %>% as.data.frame
model_df <- our_sim %>% model %>% as.data.frame
ev_with_model_params <- dplyr::right_join(model_df, ev_df, by = c("name" = "Model")) %>% tibble
f_names <- list('p-value' = "p-value",
                'beta-hat' = expression(hat(beta)))
f_labeller <- function(variable, value){return(f_names[value])}
```

```{r summary_plot, fig.dim = c(6,5) * 1.2}
ev_with_model_params %>%
  pivot_longer(cols = betahat:pvalue, names_to="eval") %>%
  mutate(eval = ifelse(eval == "pvalue", "p-value", "beta-hat")) %>%
  ggplot(aes(x = sig_sq_e, group = sig_sq_e, y = value)) +
  geom_boxplot() +
  geom_abline(aes(slope = slope, intercept = intercept), color = "blue",
              tibble(eval = c("p-value", "beta-hat"), slope = c(0,0), intercept = c(0.05, 0))) +
  xlab(expression(sigma[e]^2)) +
  # ggtitle("Corncob results for testing\nH0: Mean Gardnerella relative abundance is equal for preterm & full-term") +
  facet_wrap(~eval, scales="free_y", nrow = 2, labeller = f_labeller) +
  ylab("") +
  cowplot::theme_cowplot()
```

@@

A major advantage of this approach is that any distribution of efficiency vectors can be investigated. For example, we could specifically investigate how _Gardnerella_ being low- or high-efficiency impacts our results. We could also investigate more sophisticated distributions for the efficiencies, such as correlated efficiencies across phylogeny.

# Data setup {.appendix}

The following code was used to download and prepare the data from @callahan2017repl for analysis; it is not evaluated when this document is knit.
The first code chunk downloads the ASV abundance matrix, sample metadata, and taxonomy assignments for the vaginal samples.

```{r import_data, eval = FALSE}
library(tidyverse)
library(speedyseq)

# Download --------------------------------------------------------------------
dl_path <- here::here("notebook/_data", "callahan2017repl")
link <- "https://stacks.stanford.edu/file/druid:yb681vm1809/RepRefine_Scripts.tar.gz"
if (!dir.exists(dl_path)) {
  dir.create(dl_path)
  download.file(link, file.path(dl_path, basename(link)))
  system2("tar", args = c("-xvf", file.path(dl_path, basename(link)), "-C", dl_path))
}
load(file.path(dl_path, "RepRefine_Scripts", "input", "processed.rda"))
# This loads:
# df: the sample metadata
# st: the sequence table (in counts)
# ft: the sequence table normalized to frequencies / proportions
# tax: taxonomy assignment matrix
# 
# To confirm relationship between ft and st, run 
# all.equal(ft, st %>% apply(1, function(x) x / sum(x)) %>% t)
# 
# The taxonomy matrix has an extra column with a duplicate name of "Species"
# that is all NAs, so let's remove that
tax <- tax[, -8]
# Import into phyloseq --------------------------------------------------------
sam <- sample_data(df)
st <- otu_table(st, taxa_are_rows = FALSE)
tax <- tax_table(tax)
ps <- phyloseq(sam, st, tax)
# Check that all samples and taxa made it into the phyloseq object
stopifnot(all.equal(colnames(st), taxa_names(ps)))
stopifnot(all.equal(rownames(sam), sample_names(ps)))
# Add ASV sequences to the refseq() slot
seqs <- taxa_names(ps) %>% rlang::set_names() %>% Biostrings::DNAStringSet()
ps <- merge_phyloseq(ps, seqs)
# Rename the ASVs to simpler names (ASV1, ASV2, ...). The full sequences remain
# in `refseq(ps)`.
taxa_names(ps) <- paste0("ASV", seq(ntaxa(ps)))
# Clean up objects
rm(tax, df, sam, st, ft, seqs)
```

Next, we use munging functions from janitor and speedyseq to select the metadata of primary interest, standardize format of column names, and convert categorical variables to factors.

```{r, eval = FALSE}
ps1 <- ps %>% 
  select_sample_data(SubjectID, Cohort, Race, preterm, Outcome) %>%
  rename_with_sample_data(janitor::make_clean_names) %>%
  rename_with_tax_table(janitor::make_clean_names) %>%
  mutate_sample_data(
    across(c(subject_id, cohort, race, outcome), factor),
  )
```

Finally, we save the phyloseq object for use in the analysis.

```{r, eval = FALSE}
saveRDS(ps1, file.path(dl_path, "callahan2017repl_phyloseq_clean.rds"))
rm(ps, ps1)
```

# Session info {.appendix}

<details><summary>Click for session info</summary>
```{r, R.options = list(width = 83)}
sessioninfo::session_info()
```
</details>
